---
title: "assignment 1"
output: html_document
---

### 3.1

We chose a [dataset](https://climatedata.imf.org/datasets/b13b69ee0dde43a99c811f592af4e821/explore) that includes climate related disaster frequencies. Each observation is grouped by year and country and contains various country information (ISO codes, name, etc.), a description of the indicator, the unit, the source of the data and CTS related info. The data is in wide format, meaning that the years, within the range 1980-2024, are columns instead of values.

```{r eval=FALSE}
library(tidyverse)
```

```{r}
data = read_csv('./data/phys_risks_climate_related_distaster_frequency.csv')
glimpse(data)
```


### 3.3

```{r}
num_rows = 10
```

The `count` function allows us to group by and count all distinct entries of a specified column.

For example, we can first group by country and then apply `head` which will output the first 10 rows:

```{r}
 head(count(data, Country), n = num_rows)
```

Otherwise, if we first apply `head` and subsequently `count`, since the first 10 entries contain the same country, we will obtain a single row (with n = 10):

```{r}
count(head(data, n = num_rows), Country)
```

### 3.5

```{r}
ggplot(data) # ??
```

### 3.7


```{r}
years = as.character(1980:2024) # create a range 1980-2024 containing all col names

long_data = pivot_longer(data, years, names_to = 'Year', values_to = 'Frequency')

head(arrange(long_data, desc(Frequency)))
```





